---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# mkac


## Installation

You can install the released version of mkac from [Github](https://github.com/nxskok/mkac) with:

```{r, eval=F}
devtools::install_github("nxskok/mkac")
```

## Packages (for examples below)

```{r}
library(tidyverse)
library(mkac)
```


## Time trends

This package facilitates the investigation of two basic problems in the analysis of data collected over time:

1. Is there a time trend at all (beyond chance)?
2. If there is a time trend, how big is it?

## The Mann-Kendall correlation

The first question is attacked using the Mann-Kendall correlation, which is the Kendall correlation where the $x$-variable is time. This is a non-parametric correlation that does not assume linearity and is not damaged by outliers. See [this Wikipedia page ](https://en.wikipedia.org/wiki/Kendall_rank_correlation_coefficient) for details. It will thus detect non-linear but monotonic trends. 

The Mann-Kendall correlation is based on the idea of "concordant" and "discordant" pairs. Suppose the variable measured over time is called y, and consider two values of y measured at different time points:

- if y is smaller for the earlier time point, the pair is called **concordant**;
- if y is *larger* for the earlier time point, the pair is called **discordant**;
- if the two values of y are the same, the pair is neither concordant nor discordant, and is ignored in the calculation of the Kendall correlation.

The names reflect whether or not the pair of observations is in the same order as time; a concordant pair shows an "uphill" trend, and a discordant pair a "downhill" trend:

```{r, echo=F}
d=tribble(
  ~what, ~time, ~y,
  "concordant", 1, 10,
  "concordant", 2, 12,
  "discordant", 1, 11, 
  "discordant", 2, 9
)
ggplot(d, aes(x=time, y=y))+geom_point()+geom_line()+facet_wrap(~what)
```

The Mann-Kendall correlation is then obtained by considering all possible pairs of observations, and counting up the total number of concordant and discordant ones. The difference concordant minus discordant is then scaled to lie between -1 and 1. There are n(n-1)/2 pairs of observations, and an outlier only contributes only n-1 concordances or discordances to the total, no matter how much of an outlier it is, so outliers can have only limited influence on the Mann-Kendall correlation.

If the values of y are independent, there is standard theory that enables us to test the null hypothesis that the Mann-Kendall correlation is zero. However, data that are collected over time are often autocorrelated. [Hamed and Rao](http://resolver.scholarsportal.info/resolve/00221694/v204i1-4/182_ammttfad) obtain an approximation to the P-value of the Mann-Kendall test that applies when the observations are autocorrelated. The common case is of positive autocorrelation (when one value of y is high, the next value is more likely also to be high); the P-value for independent data is then *too low* and thus overstates the significance of the time trend. The Hamed-Rao adjustment is implemented in this package.

## Example

Consider world mean temperatures by year:

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/temperature.csv"
temp=read_csv(my_url)
temp
ggplot(temp, aes(x=year, y=temperature)) + geom_point() + geom_smooth()
```

This appears to show an upward trend, but with a lot of variability. Is that statistically significant?

```{r}
kendall_Z_adjusted(temp$temperature)
```

This says, for testing that the Mann-Kendall correlation is zero:

- Under the assumption of independent observations, the test statistic is z=11.77 with a P-value of zero. 
- Under the assumption of autocorrelated observations, the test statistic is z=4.48 with a P-value of 0.0000076.
- The autocorrelation makes the "effective sample size" 6.91 times smaller.

Because the temperatures are positively autocorrelated, the trend is strongly significant, but not as significant as it would be if the temperatures had been independent. I recommend using the adjusted P-value, but also looking at the effective sample size to see whether adjusting for autocorrelation actually made any difference. 

## Theil-Sen slope

Having found a significant trend, the next question is "how big is it"? This can be answered by looking at the Theil-Sen slope. This assumes linearity (which should be checked for reasonableness), but is not affected by outliers.

The **pairwise slope** between two observations is the difference in their values of y divided by the difference in time between them. Each pair of points has a pairwise slope, and the Theil-Sen slope is the *median* of all possible pairwise slopes. Thus an outlier observation has *no* influence on the Theil-Sen slope (it can affect n-1 of the n(n-1)/2 pairwise slopes, but even if it does, it will not affect the median of them at all).

The world mean temperatures do not have a linear trend, but the Theil-Sen slope will give us some kind of average rate of change:

```{r}
theil_sen_slope(temp$temperature)
```

This way of running `theil_sen_slope` assumes that the time points are equally spaced. If they are not, for example, you have removed some time points with missing values for the time series (see below for an example), run `theil_sen_slope` with *two* inputs. Or run it with two inputs anyway, to get the same result as above:

```{r}
with(temp, theil_sen_slope(y = temperature, x = year))
```

See below for another example of running `theil_sen_slope` this way, in the context of having removed missing values.

From our graph, the trend appears to be approximately linear up to about 1970, and approximately linear after that, but with a steeper trend. We might calculate and compare two separate Theil-Sen slopes, thus:

```{r}
temp %>% mutate(time_period=ifelse(year<=1970, "pre-1970", "post-1970")) %>% 
  rowwise() %>% 
  nest_by(time_period) %>% 
  mutate(theil_sen=theil_sen_slope(data$temperature))
```

Theil-Sen slope is very nearly *four times* as big since 1970 vs. before, and even then, appears to be increasing with time.

## Missing values and irregular time series

Your time series might contain missing values, or might be sampled at irregular intervals. The recommended way to handle a time series with missing values, at least for the purposes of Mann-Kendall or Theil-Sen, is to remove the rows with missing values, converting it into a time series sampled at irregular intervals. 

This has no impact on Mann-Kendall (a pair of points in a time series are concordant or discordant regardless of how far apart they are in time), but it does matter for Theil-Sen, since the distance apart in time affects the pairwise slope. This means using the two-input version of `theil_sen_slope` referred to above, having first removed any missing time-series values and their accompanying time points. Here is a small example:

```{r}
ts <- tribble(
  ~year, ~value,
    1,      2,
    2,      5,
    3,      4,
    4,     NA,
    5,     10
)
ts
```

First remove any rows with missing values:

```{r}
ts %>% drop_na(value) -> ts1
ts1
```

Mann-Kendall still works:

```{r}
kendall_Z_adjusted(ts1$value)
```

but now we need the two-input version of `theil_sen_slope`:

```{r}
with(ts1, theil_sen_slope(y = value, x = year))
```

If you are not familiar with `with`, this way also works, less elegantly:

```{r}
theil_sen_slope(y = ts1$value, x = ts1$year)
```
